---
title: "stats2data_application"
output:
  rmarkdown::html_vignette:
    df_print: default
vignette: >
  %\VignetteIndexEntry{stats2data_application}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 6,
  fig.height = 4,
  fig.align  = "left",
  dpi        = 300
)
```

# Overview

This vignette walks through two empirical examples based on published studies whose raw data are not publicly available:

* an **ANOVA design** with reported cell means and F-statistics (Reynolds & Besner, 2008);
* a **multiple linear regression** with reported descriptives, correlations, and regression coefficients (Bardwell et al., 2007).

Across the two examples, every exported function in `stats2data` is demonstrated:

* the optimisation functions `optim_aov()`, `optim_vec()`, `optim_mlr()`, and `parallel_optim()`;
* the S3 inspection methods `print()`, `summary()`, `get_stats()`, `get_rmse()`, and `coef()`;
* the diagnostic and comparison plots `plot_error()`, `plot_cooling()`, `plot_summary()`, and `plot()` (for parallel results);
* the GRIM helper `check_grim()`;
* the Shiny entry point `run_app()`.

Tuning hyperparameters are left at their package defaults wherever feasible.

## Installation

```{r install, echo = TRUE, eval = FALSE}
if (!requireNamespace("devtools", quietly = TRUE)) install.packages("devtools")
devtools::install_github("sebastian-lortz/stats2data")
```

## Setup

```{r load}
library(stats2data)
library(ggplot2)
library(jtools)
library(gridExtra)
set.seed(2026)
```

# 1. ANOVA module --- Reynolds & Besner (2008)

In their effort to assess the reproducibility of psychological science, the Open Science Collaboration (2015) attempted a large-scale replication of research findings. Among the replicated studies are Reynolds and Besner (2008), who reported a 2 × 2 within-subjects design (`S = 16`) measuring response times to exception words and nonwords under predictable switch and stay sequences. The original raw data are not publicly available (Open Science Collaboration, 2015). The reported cell means and F-statistics provide the targets required by `optim_aov()`. This section simulates a plausible raw dataset consistent with those targets, using the optimisation methods implemented in `stats2data`.

## 1.1 Encoding the targets

The reported summary statistics are encoded as the inputs of `optim_aov()`. The factor types, the design formula, and the data type (continuous mean response times) follow directly from the method description of the experiment.

```{r aov-targets}
S                  <- 16
levels             <- c(2, 2)
factor_type        <- c("within", "within")
formula            <- "outcome ~ Factor1 * Factor2 + Error(ID / (Factor1 * Factor2))"
integer            <- FALSE
```

In the results, `Factor1` is reported with `F(1, 15) = 30.5`. `Factor2` and the `Factor1:Factor2` interaction are reported as `F < 1`; small ordered placeholders are supplied for both. The cell means are taken directly from the manuscript:

```{r aov-fvalues}
target_f_list <- list(
  effect = c("Factor1", "Factor2", "Factor1:Factor2"),
  F_value      = c(30.5, 0.4, 0.5)
)
target_group_means <- c(543, 536, 614, 618)
```

A plausible (and deliberately wide) admissible range is derived from the reported pooled `MSE = 3070`:

```{r aov-range}
range  <- c(min(target_group_means) - floor(4 * sqrt(3070)),
            max(target_group_means) + ceiling(4 * sqrt(3070)))
range
```

## 1.2 Running the optimiser

`optim_aov()` is called with default hyperparameters (`max_iter = 1e4`, `thresh = 5e-3`, `init_temp = 1e-3`, `cooling_rate = NULL` (set automatically), `max_starts = 3`).

```{r aov-run}
result.aov <- optim_aov(
  S                  = S,
  levels             = levels,
  target_group_means = target_group_means,
  target_f_list      = target_f_list,
  factor_type        = factor_type,
  range              = range,
  formula            = formula,
  integer            = integer
)
```

## 1.3 Inspecting the result

The default `print()` method shows the design, the sample size, the requested effects, and the best objective value. The reported best error is below `thresh = 5e-3`, indicating that the optimiser has located a candidate within the user-specified threshold.

```{r aov-print}
result.aov
```

`summary()` prints the target-vs-simulated F-values, the target-vs-simulated cell means, and the corresponding RMSEs.

```{r aov-summary}
summary(result.aov)
```

`get_stats()` returns the underlying numeric components: the fitted `afex` ANOVA table, the F-vector, and the simulated cell means.

```{r aov-stats}
stats.aov <- get_stats(result.aov)
stats.aov
```

`get_rmse()` returns the F-statistic and group-mean RMSEs as a list.

```{r aov-rmse}
get_rmse(result.aov)
```

## 1.4 Diagnostics

`plot_error()` shows the simulated-annealing best-error trajectory.

```{r aov-error, fig.alt = "Error trajectory for the ANOVA optimisation."}
plot_error(result.aov)
```

`plot_cooling()` reconstructs the temperature schedule from the stored `inputs`.

```{r aov-cooling, fig.alt = "Cooling schedule for the ANOVA optimisation."}
plot_cooling(result.aov)
```

## 1.5 Target vs. simulated comparison

`plot_summary()` builds a lollipop chart of simulated minus target values. With `standardised = FALSE` (used here because one target F is below the standardisation threshold), the y-axis is on the original F-scale; with the default `standardised = TRUE`, discrepancies would be divided by the target value, with a fall-back to the unstandardised difference for near-zero targets.

```{r aov-summary-plot, fig.alt = "ANOVA target-vs-simulated lollipop chart."}
plot_summary(result.aov, standardised = FALSE)
```

## 1.6 Inspect the simulated dataset

The simulated dataset is stored at `result.aov$data` with columns `ID`, `Factor1`, `Factor2`, `outcome`.

```{r aov-data}
data.aov <- result.aov$data
head(data.aov)
```

The four cells and the four factor marginals are visualised as densities in a greyscale `theme_apa()` style.

```{r aov-cells, fig.width = 7, fig.height = 9, fig.alt = "Densities of the simulated outcome by cell and by factor marginal."}
make_panel <- function(idx, title) {
  ggplot(data.frame(value = data.aov$outcome[idx]),
         aes(x = value)) +
    geom_density(fill = "grey25", colour = "white", linewidth = 0.2) +
    labs(title = title, x = NULL, y = NULL) +
    xlim(300, 900) +
    jtools::theme_apa() +
    theme(legend.position = "none")
}

panels <- list(
  make_panel(data.aov$Factor1 == 1 & data.aov$Factor2 == 1, "Factor1=1, Factor2=1"),
  make_panel(data.aov$Factor1 == 1 & data.aov$Factor2 == 2, "Factor1=1, Factor2=2"),
  make_panel(data.aov$Factor1 == 2 & data.aov$Factor2 == 1, "Factor1=2, Factor2=1"),
  make_panel(data.aov$Factor1 == 2 & data.aov$Factor2 == 2, "Factor1=2, Factor2=2"),
  make_panel(data.aov$Factor1 == 1, "Factor1=1 marginal"),
  make_panel(data.aov$Factor1 == 2, "Factor1=2 marginal"),
  make_panel(data.aov$Factor2 == 1, "Factor2=1 marginal"),
  make_panel(data.aov$Factor2 == 2, "Factor2=2 marginal")
)

gridExtra::grid.arrange(grobs = panels, ncol = 2)
```

## 1.7 Repeated runs --- `parallel_optim()`

`parallel_optim()` repeatedly calls any `optim_*` function with the same arguments, returns the list of results, and identifies the single best run by minimum objective error. `runs = 18` parallel replications are used here; setting `cores = 6` dispatches the runs to multiple R sessions via `future::multisession`.

```{r aov-parallel, results = "hide", message = FALSE}
result.parallel_aov <- parallel_optim(
  FUN = optim_aov,
  args = list(
    S                  = S,
    levels             = levels,
    target_group_means = target_group_means,
    target_f_list      = target_f_list,
    factor_type        = factor_type,
    range              = range,
    formula            = formula,
    integer            = integer
  ),
  runs  = 18,
  cores = 6,
  seed  = 2026
)
```

### Inspect

```{r aov-par-print}
result.parallel_aov
```

```{r aov-par-summary}
summary(result.parallel_aov)
```

`get_stats()` for a `stats2data_parallel` object aggregates each numeric component returned by the per-run `get_stats()` method into a data frame with columns `mean`, `median`, `sd`, `min`, `max`:

```{r aov-par-stats}
get_stats(result.parallel_aov)
```

`get_rmse()` returns three components: `between_rmse` (per-metric run-to-run RMSE against the across-run grand mean), `target_rmse` (per-metric RMSE against the original targets), and the raw per-run vectors:

```{r aov-par-rmse}
get_rmse(result.parallel_aov)
```

### RMSE distribution

The `plot()` method for `stats2data_parallel` displays the between-run RMSE distribution alongside the deviation-from-target RMSE distribution, faceted by metric:

```{r aov-par-plot, fig.width = 8, fig.height = 4, fig.alt = "Between-run vs. target RMSE distribution across parallel ANOVA runs."}
plot(result.parallel_aov)
```

### Aggregated target-vs-simulated comparison

`plot_summary()` for a parallel result draws the same lollipop chart as for a stand-alone result; the simulated point per parameter is the across-run mean and an error bar overlays the across-run min/max (default).

```{r aov-par-summary-plot, fig.width = 8, fig.height = 4, fig.alt = "Across-run aggregated target-vs-simulated summary statistics for parallel ANOVA runs."}
plot_summary(result.parallel_aov, standardised = FALSE)
```

The single best run is accessible via `$best` and supports the same S3 methods as a stand-alone result:

```{r aov-par-best}
result.parallel_aov$best
get_rmse(result.parallel_aov$best)
```

# 2. Descriptives + MLR module --- Bardwell et al. (2007)

Bardwell et al. (2007, [https://doi.org/10.1016/j.jad.2006.06.013](https://doi.org/10.1016/j.jad.2006.06.013)) examined whether mood symptoms or apnea severity better predict daytime fatigue in patients with obstructive sleep apnea. The original raw data of Bardwell et al. (2003) are not publicly available. The simulation proceeds in three steps: marginal distributions via `optim_vec()`, joint regression structure via `optim_mlr()`, and a replication assessment via `parallel_optim()`.

## 2.1 Step 1 --- marginals via `optim_vec()`

### Targets

```{r vec-targets}
N           <- 60
target_mean <- c(48.8, 17.3, 12.6, 10.8)
names(target_mean) <- c("Apnea.1", "Apnea.2", "Depression", "Fatigue")
target_sd   <- c(27.1, 20.1, 11.3, 7.3)
integer     <- c(FALSE, FALSE, TRUE, TRUE)
range_matrix <- matrix(
  c( 15,   0,  0,  0,
    111, 80.9, 49, 28),
  nrow = 2, byrow = TRUE
)
```

### Sanity check with `check_grim()`

The GRIM test (Brown & Heathers, 2017) verifies whether a reported mean is achievable for the stated sample size and decimal precision. Each integer-valued mean is checked at one decimal place:

```{r vec-grim}
check_grim(n = N, target_mean = target_mean["Depression"], decimals = 1)
check_grim(n = N, target_mean = target_mean["Fatigue"],    decimals = 1)
```

### Run the optimiser

`sprite_prec = c(1, 1)` matches the reported one-decimal rounding precision of the target means and SDs; all other arguments are at the package defaults (`thresh = .005`, `max_iter = 5e5`, `init_temp = cooling_rate = NULL` (set automatically), `max_starts = 3`).

```{r vec-run}
result.vec <- optim_vec(
  N           = N,
  target_mean = target_mean,
  target_sd   = target_sd,
  range       = range_matrix,
  integer     = integer,
  sprite_prec = c(1, 1),
  thresh      = .05
)
```

### Inspect

```{r vec-print}
result.vec
```

```{r vec-summary}
summary(result.vec)
```

```{r vec-stats}
get_stats(result.vec)
get_rmse(result.vec)
```

### Diagnostics

`plot_error()` for a `stats2data_vec` object accepts `run` to select a per-variable trajectory; the trajectory for `Apnea.2` (the second variable) is shown:

```{r vec-error, fig.alt = "Error trajectory for the Apnea.2 marginal."}
plot_error(result.vec, run = 2)
```

```{r vec-cooling, fig.alt = "Cooling schedule for the descriptives optimisation."}
plot_cooling(result.vec)
```

### Target vs. simulated comparison

```{r vec-summary-plot, fig.alt = "Descriptives target-vs-simulated lollipop chart."}
plot_summary(result.vec, standardised = FALSE)
```

### Inspect the simulated dataset

The four marginals are shown together. `Apnea.1` and `Apnea.2` are continuous; `Depression` and `Fatigue` are integer-valued (`integer = c(FALSE, FALSE, TRUE, TRUE)`), so the latter two are shown as `binwidth = 1` histograms scaled to density to remain visually comparable to the densities of the continuous variables.

```{r vec-data, fig.width = 8, fig.height = 6, fig.alt = "Marginal distributions of the four simulated variables."}
data.vec <- result.vec$data
head(data.vec)

cont_vars <- c("Apnea.1", "Apnea.2")
int_vars  <- c("Depression", "Fatigue")

ggdata.vec <- tidyr::pivot_longer(
  data.vec,
  cols      = c(Apnea.1, Apnea.2, Depression, Fatigue),
  names_to  = "variable",
  values_to = "value"
)
ggdata.vec$variable <- factor(
  ggdata.vec$variable,
  levels = c(cont_vars, int_vars)
)
ggdata.vec$kind <- ifelse(ggdata.vec$variable %in% int_vars,
                          "integer", "continuous")

ggplot(ggdata.vec, aes(x = value)) +
  geom_density(
    data  = subset(ggdata.vec, kind == "continuous"),
    fill  = "gray60", color = "gray20", alpha = 0.5
  ) +
  geom_histogram(
    data     = subset(ggdata.vec, kind == "integer"),
    aes(y    = after_stat(density)),
    binwidth = 1,
    fill     = "gray60", color = "gray20"
  ) +
  facet_wrap(~ variable, scales = "free", nrow = 2) +
  labs(x = NULL, y = "Density") +
  theme_minimal(base_size = 13)
```

## 2.2 Step 2 --- regression structure via `optim_mlr()`

### Targets

The reported regression coefficients (intercept first), their standard errors, and the upper-triangular correlation matrix (column-wise, excluding the diagonal) are encoded as numeric vectors. `NA` entries flag unreported values that are excluded from the objective.

```{r lm-targets}
target_reg <- c(4.020, 0.023, 0.008, 0.438)
names(target_reg) <- c("(Intercept)", "Apnea.1", "Apnea.2", "Depression")
target_se   <- c(NA, 0.034, 0.048, 0.066)
target_cor  <- c(NA, NA, NA, 0.11, 0.20, 0.68)
reg_equation <- "Fatigue ~ Apnea.1 + Apnea.2 + Depression"
```

### Run the optimiser

A first call uses package defaults except for a stricter `thresh = .0005` and a starting `weight = .5` for the regression-vs.-correlation balance.

```{r lm-run}
result.mlr <- optim_mlr(
  N            = N,
  target_mean  = target_mean,
  target_sd    = target_sd,
  range        = range_matrix,
  integer      = integer,
  reg_equation = reg_equation,
  target_cor   = target_cor,
  target_reg   = target_reg,
  target_se    = target_se,
  thresh       = .0005,
  weight       = .5
)
```

### Inspect

```{r lm-print}
result.mlr
```

```{r lm-summary}
summary(result.mlr)
```

`get_stats()` returns the fitted `lm`, its coefficients and standard errors, the upper-triangular correlation vector, and the column-wise means and SDs:

```{r lm-stats}
stats.mlr <- get_stats(result.mlr)
stats.mlr$reg
stats.mlr$se
stats.mlr$cor
```

`coef()` is a convenience wrapper that re-fits the regression and returns the coefficient vector:

```{r lm-coef}
coef(result.mlr)
```

```{r lm-rmse}
get_rmse(result.mlr)
```

### Diagnostics

`plot_error()` shows the total-objective trajectory:

```{r lm-error, fig.alt = "Error trajectory for the MLR optimisation."}
plot_error(result.mlr, first_iter = 1000)
```

For `stats2data_mlr` objects, `plot_error()` exposes a `ratio = TRUE` switch that plots the per-iteration correlation-to-regression error ratio together with mean, median, and final reference lines:

```{r lm-error-ratio, fig.alt = "Cor/Reg error ratio trajectory for MLR."}
plot_error(result.mlr, ratio = TRUE)
```

The trajectory shows that the regression component dominates the objective. To rebalance the two error components, the optimisation is repeated with a smaller `weight = .05`:

```{r lm-rerun, fig.alt = "Weight-tuned MLR error ratio."}
result.mlr <- optim_mlr(
  N            = N,
  target_mean  = target_mean,
  target_sd    = target_sd,
  range        = range_matrix,
  integer      = integer,
  reg_equation = reg_equation,
  target_cor   = target_cor,
  target_reg   = target_reg,
  target_se    = target_se,
  thresh       = .0005,
  weight       = .05
)
plot_error(result.mlr, ratio = TRUE)
```

The error ratio is now closer to one and the best objective value is lower than in the first call.

### Target vs. simulated comparison

```{r lm-summary-plot, fig.alt = "MLR target-vs-simulated lollipop chart."}
plot_summary(result.mlr, standardised = FALSE)
```

### Inspect the simulated dataset

```{r lm-data}
data.mlr <- result.mlr$data
head(data.mlr)
```

`plot_partial_regression()` returns one partial-regression ggplot per predictor, each annotated with the partial slope and the residual SD:

```{r lm-partial, fig.width = 11, fig.height = 8, fig.alt = "Partial-regression plots for the MLR fit."}
partial.plots <- plot_partial_regression(lm(reg_equation, data = data.mlr))
gridExtra::grid.arrange(grobs = partial.plots, ncol = 2)
```

## 2.3 Step 3 --- repeated runs via `parallel_optim()`

`parallel_optim()` repeatedly calls any `optim_*` function with the same arguments, returns the list of results, and identifies the single best run by minimum objective error. `runs = 18` parallel replications are used; setting `cores = 6` dispatches the runs to multiple R sessions via `future::multisession`.

```{r lm-parallel, results = "hide", message = FALSE}
result.parallel <- parallel_optim(
  FUN = optim_mlr,
  args = list(
    N            = N,
    target_mean  = target_mean,
    target_sd    = target_sd,
    range        = range_matrix,
    integer      = integer,
    reg_equation = reg_equation,
    target_cor   = target_cor,
    target_reg   = target_reg,
    target_se    = target_se,
    thresh       = .0005,
    weight       = .05
  ),
  runs  = 18,
  cores = 6,
  seed  = 2026
)
```

### Inspect

```{r par-print}
result.parallel
```

```{r par-summary}
summary(result.parallel)
```

```{r par-stats}
get_stats(result.parallel)
```

```{r par-rmse}
get_rmse(result.parallel)
```

### RMSE distribution

```{r par-plot, fig.width = 8, fig.height = 4, fig.alt = "Between-run vs. target RMSE distribution across parallel MLR runs."}
plot(result.parallel)
```

### Aggregated target-vs-simulated comparison

```{r par-summary-plot, fig.width = 8, fig.height = 4, fig.alt = "Across-run aggregated target-vs-simulated lollipop chart for parallel MLR runs."}
plot_summary(result.parallel, standardised = FALSE)
```

The single best run is accessible via `$best` and supports the same S3 methods as a stand-alone result:

```{r par-best}
result.parallel$best
get_rmse(result.parallel$best)
```

# 3. Shiny interface

A point-and-click interface to all three modules is launched via:

```{r app, eval = FALSE}
run_app()
```

# References
- Bardwell, W. A., Moore, P., Ancoli-Israel, S., & Dimsdale, J. E. (2003). Fatigue in obstructive sleep apnea: driven by depressive symptoms instead of apnea severity?. The American journal of psychiatry, 160(2), 350–355. https://doi.org/10.1176/appi.ajp.160.2.350
- Bardwell, W. A., Ancoli-Israel, S., & Dimsdale, J. E. (2007). Comparison of the effects of depressive symptoms and apnea severity on fatigue in patients with obstructive sleep apnea: a replication study. Journal of affective disorders, 97(1-3), 181–186. https://doi.org/10.1016/j.jad.2006.06.013
- Brown, N. J. L., & Heathers, J. A. J. (2017). The GRIM Test: A Simple Technique Detects Numerous Anomalies in the Reporting of Results in Psychology. Social Psychological and Personality Science, 8(4), 363–369. https://doi.org/10.1177/1948550616673876
- Open Science Collaboration. (2015). Estimating the reproducibility of psychological science. Science, 349(6251), aac4716. https://doi.org/10.1126/science.aac4716
- Reynolds, M., & Besner, D. (2008). Contextual effects on reading aloud: Evidence for pathway control. Journal of Experimental Psychology: Learning, Memory, and Cognition, 34(1), 50–64. https://doi.org/10.1037/0278-7393.34.1.50
